---
title: "Homework 5 - Advanced Data in R"
format: docx
---

```{r}
#| include: false
library(gapminder)
data(gapminder)
```

Fill out [this Quarto document](https://github.com/BIFX552-22/CourseInfo/hw/05advanced_data.qmd) while completing the lesson on graphics, vectorization and functions in R.

### [Writing Data](https://swcarpentry.github.io/r-novice-gapminder/11-writing-data/index.html)

::: callout-tip
## Key Points

-   Save plots from RStudio using the 'Export' button.

-   Use `write.table` to save tabular data.
:::

#### Challenge 1

Modify the code below to print a second page in the pdf, showing a facet plot (hint: use `facet_grid`) of the same data with one panel per continent.

```{r}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)
ggplot(gapminder, aes(year, lifeExp, color=country)) +
  geom_line() +
  theme(legend.position = "none")

# You then have to make sure to turn off the pdf device!

dev.off()
```

#### Challenge 2

Write a data-cleaning script file that subsets the gapminder data to include only data points collected since 1990.

Use this script to write out the new subset to a file in the `cleaned-data/` directory.

```{r}
# insert code here
```

### [Data frames with `dplyr`](https://swcarpentry.github.io/r-novice-gapminder/13-dplyr/index.html)

::: callout-tip
## Key Points

-   Use the `dplyr` package to manipulate data frames.
-   Use `select()` to choose variables from a data frame.
-   Use `filter()` to choose data based on values.
-   Use `group_by()` and `summarize()` to work with subsets of data.
-   Use `mutate()` to create new variables.
:::

#### Challenge 1

Write a single command (which can span multiple lines and includes pipes) that will produce a data frame that has the African values for `lifeExp`, `country` and `year`, but not for other Continents. How many rows does your data frame have and why?

```{r}
# insert code here
```

#### Challenge 2

Calculate the average life expectancy per country. Which has the longest average life expectancy and which has the shortest average life expectancy?

```{r}
# insert code here
```

#### Challenge 3

Calculate the average life expectancy in 2002 of 2 randomly selected countries for each continent. Then arrange the continent names in reverse order. **Hint:** Use the `dplyr` functions `arrange()` and `sample_n()`, they have similar syntax to other dplyr functions.

```{r}
# insert code here
```

### [Data frames with `tidyr`](https://swcarpentry.github.io/r-novice-gapminder/14-tidyr/index.html)

::: callout-tip
## Key Points

-   Use the `tidyr` package to change the layout of data frames.

-   Use `pivot_longer()` to go from wide to longer layout.

-   Use `pivot_wider()` to go from long to wider layout.
:::

#### Challenge 1

Is gapminder a purely long, purely wide, or some intermediate format?

#### Challenge 2

Using `gap_long`, calculate the mean life expectancy, population, and gdpPercap for each continent. **Hint:** use the `group_by()` and `summarize()` functions we learned in the `dplyr` lesson

```{r}
# insert code here
```

#### Challenge 3

Create a `gap_ludicrously_wide` format data by pivoting over countries, year and the 3 metrics? **Hint** this new data frame should only have 5 rows.

```{r}
# insert code here
```
